Identification of nonsense-mediated decay inhibitors that alter the tumor immune landscape

Version 1

Abstract

Summary

Despite exciting developments in cancer immunotherapy, its broad application is limited by the paucity of targetable antigens on the tumor cell surface. As an intrinsic cellular pathway, nonsense- mediated decay (NMD) conceals neoantigens through the destruction of the RNA products from genes harboring truncating mutations. We developed and conducted a high throughput screen, based on the ratiometric analysis of transcripts, to identify critical mediators of NMD. This screen revealed disruption of kinase SMG1’s phosphorylation of UPF1 as a potent disruptor of NMD. This led us to design a novel SMG1 inhibitor, KVS0001, that elevates the expression of transcripts and proteins resulting from truncating mutations in vivo and in vitro. Most importantly, KVS0001 concomitantly increased the presentation of immune-targetable HLA class I-associated peptides from NMD-downregulated proteins on the surface of cancer cells. KVS0001 provides new opportunities for studying NMD and the diseases in which NMD plays a role, including cancer and inherited diseases.

Introduction

Despite success with cancer immunotherapies, approved immunotherapies are not available for the majority of cancer patients and only a minority of treated patients realize a durable response1–4. Current studies largely focus on discovering new agents and identifying patients most likely to benefit from existing immunotherapies5. While many studies correlate tumor insertion and deletion (indel) mutation load with immunotherapeutic response, not all tumors with high indel mutational loads respond to checkpoint inhibitors6–13.

The typical adult solid tumor contains a median of 54 coding somatic nucleotide variants, many of which have the potential to create novel neoantigens or Mutation Associated NeoAntigens (MANA)14–16. Approximately 5% of mutations are insertions/deletions (indels) or splice site changes that alter the open reading frame (ORF) of the transcript14–16. This subset of mutations is of particular interest because they can result in proteins and derived peptides that are foreign to a host’s healthy cells, giving rise to neoantigens17–19. In normal cells, nonsense-mediated decay (NMD) plays an important role in messenger RNA (mRNA) quality control, as well as normal gene expression20,21. In cancer cells, however, NMD may aid immuno-evasion by eliminating RNA transcripts coming from genes that carry truncating mutants19. This prevents translation and presentation of peptides from these proteins on MHC class I complexes, rendering them invisible to immune cells22. Previous work has shown knockdown of the NMD pathway with siRNA enhances the anti-tumor immune response18,23. Analysis of cell lines with loss of the UPF1 RNA helicase and ATPase gene (UPF1), a key mediator of NMD degradation, also showed increased levels of aberrant transcripts and mutant proteins in the alleles targeted by NMD24. This is reminiscent of the pharmacological modulation of splicing, which has also been shown to increase the number of neoantigens present on the cancer cell surface due to similar underlying mechanisms25. Despite early studies showing little toxicity with NMD inhibition, there is as of yet no reports of a specific chemical inhibitor of the pathway with good bioavailability26–30.

In this work we create a cell-based high throughput assay to query the effects of a curated library of small molecules on NMD function. From this screen, we identify a lead compound capable of inhibiting NMD and ascertain its protein target as nonsense mediated mRNA decay associated PI3K related kinase (SMG1), a kinase that is a critical mediator of NMD. As the lead compound produced unacceptable toxicity in animal models, we then designed a specific and bioavailable small molecule inhibitor of SMG1 (KVS0001) that is well tolerated in vivo. We then demonstrate that targeted inhibition of SMG1 by KVS0001 leads to the presentation of novel neoantigens identifiable by T cells leading to tumor growth inhibition in vitro and in vivo.

Results

Development of a high throughput assay to identify NMD inhibitors

To develop an assay to find NMD inhibitors, we identified isogenic cell lines with out-of-frame indel mutations, hereinafter referred to as truncating mutations, targeted by NMD activity. We previously reported a panel of non-cancerous cell lines in which 19 common tumor suppressor genes were inactivated using the Clustered Regularly Interspaced Short Palindromic Repeats (CRISPR)-Cas9 system31. Through evaluation of this panel, we discovered two genes, Stromal Antigen 2 (STAG2) and Tumor Protein p53 (TP53), which did not express their expected proteins when assessed by western blots (STAG2) or immunohistochemistry (p53) (Figure S1). Because the inactivation of these two genes was the result of frame-shift mutations, we suspected that the absence of the proteins was due to NMD. This suspicion was supported by whole transcriptome RNA-sequencing (Figure S2)31. Notably we saw an average decrease of STAG2 RNA transcripts by 20-fold and TP53 by 6-fold relative to their respective wild-type transcripts in the parental cell lines.

We selected two STAG2 knockout clones (clone 2 and 8) and one TP53 knockout clone (clone 221) derived from the Retinal Pigmented Epithelial (RPE1) cell line to design a high-throughput screen (HTS). Treatment with the canonical NMD inhibitor emetine demonstrated up to a 60-fold recovery of mutant RNA transcript expression, establishing this panel as appropriate for HTS (Figure S3)32.

We then designed a Next Generation Sequencing (NGS) assay for NMD (Figure 1A). We mixed the three cell lines in equal proportions and plated the mixture in 96-wells plates followed by treatment with one compound (drug) per well. We determined NMD inhibition efficacy by comparing the wild-type and truncating mutant transcript expression levels in a ratiometric manner. Specifically, the wild-type sequences of the reciprocally knocked out clone (i.e., wild-type STAG2 sequence from the TP53 knockout clones, and wild-type TP53 sequence from the STAG2 knockout clones) served as internal references, providing a ratiometric assay of mutant to wild-type transcript abundance. This ratiometric assay minimized confounders introduced by nonspecific transcriptional activators or generally toxic agents.

LY3023414 is a small molecule capable of increasing transcription of NMD targets.

A) Schematic of HTS used to identify inhibitors of NMD. Mutant transcripts are represented by a smaller length in the cartoon for illustrative purposes only. B) Mutant RNA reads relative to wild-type reads for the top 8 hits from the HTS. The dotted line represents the minimum fraction required to be considered a hit (> 5 standard deviations above DMSO control). Full screen results are presented in Figure S4. C) Targeted RNA sequencing results of isogenic RPTec knockout clones treated with the 8 best hits from the HTS at 10 µM. The dotted line represents a relative RNA expression level of 1, equal to that of DMSO treated wells. Data for ceritinib, which did not validate on any line, is presented only in Figure S8. D) TP53 western blot on RPE TP53 224, containing a homozygous TP53 mutation, using the four hit compounds that validated in RPTec isogenic lines. E) Western blot analysis of full-length TP53α and isoform TP53β after treatment with two NMD inhibitor lead candidates. TP53β (expression known to be controlled by NMD) as well as mutant TP53 are prominently induced by LY3023414 whereas full-length is not. Note that RPE TP53 223 is a heterozygous knockout clone with one near wild-type allele whereas RPTec TP53 588 contains a homozygous TP53 indel mutation. F) qPCR showing LY3023414 treatment causes increased expression of the NMD controlled alternative transcript for TP53, TP53β, in parent cell lines for RPE1 and RPTec. Significance determined by Student’s t-test. Unless indicated otherwise cells were exposed to 5 µM of test compound for 16 hours.

The use of three cell lines with different truncating mutations from two different target genes minimized the possibility that drugs identified in the HTS were cell-line clone or mutation-specific. Note that the use of these cell lines, carefully mixed, banked, and preserved, did not substantially increase the amount of time or work required to screen a single cell line. A combination of well and plate barcodes allowed the pooling and scoring of over 1920 assays in a single NGS lane (see methods).

Execution of a high throughput screen to identify NMD inhibitors

Previous human clinical trials suggest that off-target toxicity at doses required for NMD inhibition makes emetine and other well-known NMD inhibitors unsuitable for human use33–37. We performed a high throughput screen (HTS) to identify more specific NMD inhibitors by treating the isogenic cell line panel described above with a commercially available library consisting of 2,658 FDA- approved or in late-phase clinical trial small molecules and natural products (Table S1). After purifying RNA from cells 16 hours post-treatment, we scored NMD inhibition using the strategy described in Figure 1A and Figure S4. Predictably, emetine increased the relative expression of mutant to wild-type transcripts by 3 to 4 fold on average (Figure S5). Eight compounds (0.3% of the library) increased the ratiometric mutant transcript fraction more than five standard deviations above the DMSO controls in all three cell lines (Figure 1B). This hit threshold was chosen as it was the minimum required to ensure no false positives were observed in the DMSO controls. One of these eight compounds, anisomycin, is a known inhibitor of protein synthesis and commonly used NMD inhibitor for in vitro studies, providing independent validation of the screen32. The other seven compounds increased mutant RNA transcript levels 5 to 10-fold relative to untreated cells.

LY3023414 inhibits NMD and causes re-expression of mutant RNA and protein

To validate the eight hit compounds described above, we tested their effects in additional lines with mutations targeted by NMD. First, we assessed them on isogenic STAG2 and TP53 knockouts in RPtec cells, another non-cancerous cell line (Figures S6A, S6B, and S7)31. Four of the original eight hit compounds increased mutant RNA expression in a dose-dependent manner (Figures 1C and S8). Next, we examined the effects of these four compounds on additional RPE1 TP53 knockout cell lines with different mutations predicted to generate truncated TP53 proteins (Figures S6C and S7). While treatment with all four potential NMD inhibitors restored expression of the truncated mutant TP53 proteins, two of them (LY3023414 and halofuginone) did so most robustly (Figures 1D and 1E). Additionally, TP53 has an isoform, TP53β, whose expression is known to be controlled by the NMD pathway38. Using cell lines with intact TP53β isoform transcripts, we observed an increase in the TP53β isoform in cell lines treated with both LY3023414 and halofuginone (Figure 1E middle and left column). Quantitative real time PCR (qPCR) of both the full-length TP53 (TP53α) and NMD sensitive (TP53β) transcripts in parental RPE1 and RPTec cells treated with LY3023414 or halofuginone confirmed the selective upregulation of TP53β, but not TP53α, transcripts (Figure 1F). Based on a consistently stronger effect of LY3023414 over halofuginone across multiple isogenic cell lines and assays, LY3023414 was chosen to be the initial lead compound for further studies involving NMD inhibition.

LY3023414 increases expression of NMD repressed mutant RNA transcripts and proteins in vitro and in vivo.

To evaluate whether LY3023414 could relieve NMD repression of naturally occurring heterozygous mutant transcripts, we chose the NCI-H358 and LS180 cancer cell lines. Treatment with LY3023414 followed by whole transcriptome RNA sequencing revealed increased expression of the mutant allele in 42% and 67% of heterozygous, out-of-frame, indel mutations in these two lines (Figure 2A). A third of the nonsense mutations in NCI-H358 were also “recovered” (i.e., mutant transcripts increased relative to wildtype transcripts) after treatment with LY3023414 (Figure 2A middle). In LS180 there was only one nonsense mutation meeting the required coverage, so it could not be evaluated in depth. Single base pair substitutions not resulting in stop codons were not affected by LY3023414 in either line (Figure 2A right).

Inhibiting NMD in cancer cells increases broad expression of truncated gene mRNA and protein.

A) Mutant transcript recovery rates for genes containing heterozygous indel mutations based on RNA-sequencing results in cell lines treated with 5µM LY3023414 for 16 hours. Strict inclusion criteria were used, such that only mutations with sufficient sequencing coverage are shown (see methods).Recovery is defined as at least 2-fold increase over DMSO treatment. B) Targeted high coverage RNA- sequencing confirms recovery of mutant transcript levels in NCI-H358 and LS180 cancer cell lines treated with LY3023414. RNF43 and DROSHA contain common heterozygous SNPs and the mutant allele refers to the non-reference genome allele. Error bars indicate 95% confidence limits. C) Western blot analyses of NCI-H358 cells showing mutant and wild-type protein levels in EXOC1 and D) SPTAN1 with and without 5 µM LY3023414 treatment. The black arrow indicates the expected size of the mutant protein. The C-terminal SPTAN1 antibody is downstream of the out-of-frame indel mutation and is not expected to identify the mutant allele. E) Fold change in the number of mutant RNA transcripts from deep-targeted RNA sequencing of heterozygous mutated genes in NCI-H358 and LS180 xenografts treated by oral gavage with 60 mg/kg LY3023414 assayed 16 hours post-treatment. Student’s t-test for target genes are all p &lt; 0.05, while the null hypothesis holds for RNF43 (common SNP).

The whole transcriptome sequencing data was confirmed by targeted deep RNA sequencing of two naturally occurring heterozygous truncating mutations from each cell line. Increased expression of the truncated mutant alleles relative to the wild-type alleles of the EXOC1 and SPTAN1 genes was observed in NCI-H358 cells after treatment with LY3023414 (Figure 2B left). Likewise, increased expression of HNRNPA3 and LRPPRC mutant transcripts was observed in LS180 after LY3023414 treatment (Figure 2B right). We noted no significant effects on two coding region heterozygous single nucleotide polymorphisms (SNP’s) after LY3023414 treatment in the genes RNF43 or DROSHA in either cell line (Figure 2B). To substantiate these transcriptomic effects at the protein level, we treated NCI- H358 cells for 24 hours with LY3023414. Western blotting revealed truncated mutant protein for both EXOC1 and SPTAN1 after treatment with LY3023414, whereas the full-length protein levels remained present and unchanged in both the treated and control (DMSO) samples (Figure 2C and 2D left)39. An antibody that recognizes the C-terminus of SPTAN1, encoded downstream of the frameshift mutation, did not detect mutant protein, as expected (Figure 2D right).

LY3023414 was originally developed as a phosphatidylinositol 3-kinase (PI3K) inhibitor with activity against AKT Serine/Threonine Kinase 1 (AKT1) and Mammalian Target of Rapamycin (mTOR)40. It was tested in a number of clinical trials and has a well-known pharmacokinetic profile in vivo41–45. To determine whether LY3023414 affects NMD in vivo, we established xenograft tumors of both NCI-H358 and LS180 in nude mice. Treatment of these mice with a single oral dose of 60 mg/kg LY3023414 led to a significant increase in the expression of mutant RNA transcripts relative to wild-type transcripts 16 hours later (Figure 2E). The RNF43 gene, which harbors a coding region heterozygous SNP, served as a control (Figure 2E). Severe drug-associated toxicity, including weight loss bordering on cachexia and near total inactivity, precluded the longer-term dosing required for anti-tumorigenic effects of LY3023414 in both BALB/c and C57BL/6N immunocompetent mouse strains.

The kinase SMG1 is the target for NMD inhibition by LY3023414

To investigate the mechanism of NMD inhibition by LY3023414, we evaluated the six kinases with the highest reported inhibition by LY302341440. siRNA-mediated knockdown of each of these kinases in NCI-H358 and LS180 cancer cell lines was performed for this purpose (Figure S9). Only knockdown of SMG1 resulted in significant changes in the amount of truncating mutant transcript relative to wild-type transcript in all four genes evaluated (Figure 3A top). RNF43 and DROSHA contain heterozygous coding region SNP’s and served as controls in these experiments and as expected showed no changes despite siRNA treatment (Figure 3A bottom).

Novel NMD inhibitor KVS0001 is SMG1 specific and induces expression of NMD targeted genes in vitro and in vivo.

A) Fraction of mutant allele transcripts in genes with heterozygous indels previously established in this study as sensitive to NMD inhibition. Results show mutant levels after siRNA treatment targeting kinases inhibited by LY3023414. RNF43 and DROSHA are common heterozygous SNPs (shaded gray) and serve as negative controls. B) Fraction of mutant allele transcripts in genes with truncating mutations known to be sensitive to NMD inhibition after siRNA treatment with siUPF1 or non-targeting siRNA. Data from deep-targeted RNA-sequencing. C) Structure of novel NMD inhibitor KVS0001. D) Targeted RNA-sequencing on three genes with heterozygous, out-of-frame, indel mutations in LS180 cancer cells treated in a dose-response with KVS0001 or SMG1i-11. RNF43 serves as a control (common heterozygous SNP) and the mutant allele refers to the non-reference genome allele. E) Western blot of EXOC1 protein in NCI-H358 cells treated with 5µM novel inhibitor KVS0001, LY3023414 or SMG1i-11 for 24 hours. F) Western blot of phosphorylated UPF1 on three cell lines treated with KVS0001, SMG1i-11, or DMSO. Note that total UPF1 and p-UPF1 were run on different gels, loading controls correspond to indicated gel. G) Fold change in the number of mutant allele transcripts measured by targeted RNA-Seq in genes containing heterozygous out-of-frame indel mutations in NCI-H358 or H) LS180 subcutaneous xenografts in bilateral flanks of nude mice. Mice were treated once with IP injection of vehicle or 30mg/kg KVS0001 and tumors harvested 16hrs post IP treatment. All genes shown contain heterozygous out-of-frame truncating mutations except RNF43 and DROSHA which serve as controls (contain heterozygous SNP’s).

SMG1 is known to regulate the NMD pathway by activating UPF1, an enzyme with RNA helicase and ATPase activity46,47. We therefore knocked down UPF1 with siRNA and found that it restored expression of the NMD-downregulated transcripts, at levels similar to those observed after the knock-down of SMG1 (Figure 3B). Additionally, treatment of NCI-H358 and LS180 cancer cell lines with a previously reported SMG1 specific small molecule inhibitor, SMG1i-11 resulted in specific increases in the transcripts and proteins from genes with truncating mutations, just as did LY3023414 (Figures S10 and S11)30. Although SMG1i-11 displayed considerable SMG1 specificity and marked inhibition of NMD, it was highly insoluble. We were unable to find a vehicle to administer it in vivo, despite numerous attempts at various formulations and administration routes. This may explain why SMG1i-11 has been demonstrated to be an effective SMG1 inhibitor in vitro, but no peer-reviewed reports of its in vivo activity have been reported to date30,48,49.

Development of an improved NMD inhibitor targeting SMG1

In view of the problems with LY3023414 and SMG1i-11, we sought to develop a new SMG1 inhibitor based on the cryo-electron microscopy structure of the binding pocket of SMG150,51. We attempted the synthesis of eleven compounds (KVS0001 to KVS0011) and tested for bioavailability and preservation of target specificity. Among these, KVS0001 stood out due to its solubility while preserving SMG1 inhibitory activity (Table S2, Figures 3C and S12). Mass spectrometry-based assays showed that KVS0001 inhibits SMG1 protein more than any of the other 246 protein or lipid kinases tested at concentrations from 10nM to 1µM (Table S3 and Figure S13)52. Noteworthy off-target kinase inhibition was not observed until doses of 1µM and above. Experiments using NCI-H358 and LS180 cells showed that KVS0001 is bioactive in the nanomolar range and subverts the NMD-mediated down-regulation of truncating mutant transcripts and proteins (Figures 3D, 3E, S14, S15). Inhibition at concentrations as low as 600nM led to equal expression of wild-type and mutant transcripts, suggesting near total blockade of the NMD pathway (Figures 3D, S14). Western blotting showed that KVS0001 substantially decreases the amount of phosphorylated UPF1, the downstream target of SMG1 kinase activity, in three different cell lines (Figure 3F). Finally, KVS0001 treatment of NCI-H358 and LS180-derived xenograft tumors in nude mice resulted in significant increases in transcript levels in each of six tested endogenous genes with truncating mutations, while having had no measurable effects on genes containing heterozygous coding region SNP’s (Figures 3G and 3H).

NMD inhibition with KVS0001 causes MHC class I display of hidden neoantigens

Cancer cells may evade immune surveillance by downregulating genes with truncating mutations as a result of NMD19,53. Indeed, previous studies with non-specific or toxic NMD inhibitors have shown an increase in selected antigens from tumor-specific mutations when NMD is inhibited54. We therefore investigated whether KVS0001 could similarly alter the cell surface presentation of proteins from genes harboring truncating mutations in NCI-H358 and LS180 cells. Using quantitative HPLC-mass spectrometry (MS) we observed a striking (45 to 90 fold) increase in the EXOC1 and RAB14 derived neoantigens, and a significant (2 fold) increase in the ZDHHC16 derived neoantigen (Figures 4A, 4B, S16, S17)55. This is consistent with the re-expression of mutant transcript and protein shown previously in this study (Figures 3D and 3E). These three peptides were chosen based on an in-silico review of potentially presented peptides and a preliminary experiment that looked at qualitative (present or not present) presentation of the predicted peptides56,57.

KVS0001 treatment induces targetable cell surface presentation of peptides known to be downregulated by NMD.

A) MHC class I HLA presentation of mutant specific peptide sequences from NCI-H358 and B) LS180 cells by quantitative HPLC-Mass Spectrometry. The gene name, type of mutation (in parenthesis) and presented peptide are shown on the y-axis for each gene. Colors indicate different ions. C) TP53 gene structure and mutant DNA sequence for NCI-H716 and NCI-H2228 cancer cell lines, both contain a homozygous splice site mutation in TP53. Capital letters represent exonic sequence; lowercase letters represent intronic sequence. DNA mutation reflected by gold bases. D) Western blot against TP53 in the presence or absence of NMD inhibitor in NCI-H716_A24 and NCI- H2228 cell lines. NCI-H2228 has an expected size of 46.6 kDa and NCI-H716 of 34.7 kDa. E) IFN-γ levels over baseline based on ELISA in a co-culture assay with NCI-H716_A24 and NCI-H2228 cells, NMD inhibitor, human CD3+ T-cells, and bispecific antibody for TP53 and CD3. Chemotherapy (5- Fluorouracil) is shown as a control. F) Cell killing based on luciferase levels in a co-culture assay in NCI- H716 cells with and without A24 expression, treated with TP53-CD3 bispecific antibody, NMD inhibitor and human CD3+ T-cells.

Targetable peptide presentation occurs with NMD inhibition by KVS0001

To test whether cancer cell neoantigens presented as a result of NMD inhibition could be targeted, we evaluated two cancer cell lines, NCI-H716 and NCI-H2228. Both lines contain homozygous truncating mutations in TP53 which produce transcripts down-regulated by NMD, with each exhibiting RNA levels less than 20% of the median level of expression among 675 cancer cell lines (Figures 4C and S18)58,59. Treatment with KVS0001 increased the expression of TP53 in both lines, while the commonly used therapeutic agents 5-fluorouracil and etoposide did not affect TP53 protein abundance (Figures 4D and S19).

To determine whether this disruption of NMD repression is targetable by T-cells, we developed a bispecific antibody (KVS-BI043) that recognizes a peptide-HLA complex on one end and CD3 on the other end. CD3 is expressed only on T-cells, and this bispecific antibody functions as a T-cell engager, linking target cells to cytotoxic T-cells, which then kill the targets60,61. KVS-BI043 recognizes a ten amino acid peptide (residues 125 to 134 of TP53) bound to HLA-A24. NCI-H2228 naturally express A24 whereas NCI-H716 cells were engineered to express A24 using a retrovirus (NCI-H716_A24). Treatment of NCI-H716_A24 or NCI-H2228 cells with KVS0001 and the KVS-BI043 bispecific antibody in the presence of normal T-cells caused a significant increase in interferon (IFN-γ) release (Figure 4E). We observed no changes in IFN-γ levels in the absence of KVS-BI043 or normal T-cells (Figure 4E grey boxed lanes). Most importantly, treatment of NCI-H716_A24 with KVS0001 in the presence of KVS- BI043 and T-cells led to significant killing of the target cancer cells, which was also not observed in the absence of T-cells or the absence of HLA-24 in the target cells (Figure 4F). NCI-H2228 was assessed for killing but expressed too much TP53 at baseline (Figure 4D left most lane) and thus displayed substantial killing even in the DMSO controls.

Tumor growth is slowed in mice treated with KVS0001

Finally, we investigated whether KVS0001 could impact tumor growth in syngeneic models in which the native immune system might play a role. For this purpose, we first used murine RENCA (renal cancer) and LLC (lung cancer) as they are known to have a relatively large number of out-of-frame indel mutations (Table S4). Although murine and human SMG1 are highly related (98% at the amino acid level), it was important to show that KVS0001 could actually inhibit NMD in murine cells. For this purpose, we tested eight genes in LLC, and four in RENCA, which contained out-of-frame indel mutations potentially targeted by NMD. We also assessed the expression of three genes without any mutations known to have their normal expression controlled by NMD62. We found that six of the twelve truncating mutation-containing genes and five of the six expression controlled by NMD genes had significantly increased RNA following in vitro treatment with KVS0001 (Figure 5A and 5B).

I n vivo treatment of murine tumors with KVS0001 yield differential tumor growth compared with vehicle treatment.

A) Fold change in RNA transcript levels in LLC or B) RENCA cells treated in-vitro with 5µM of NMD inhibitor KVS0001 or DMSO. Orange bars indicate genes with homozygous indel mutations potentially targeted by NMD. Purple bars show genes with no mutations but that are known to have their normal transcription levels controlled by NMD. Green bar is a control gene that should not change with treatment. The dotted line shows relative expression of DMSO treatment (equal to 1). C) Treatment schedule for syngeneic tumor mouse experiments. D) Average tumor size of LLC (left) and RENCA (right) syngeneic tumors in immune-competent mice (n=8) treated with 30 mg/kg KVS0001 or vehicle control. Difference is statistically significant after day 10 based on one-way ANOVA with Dunnett’s test p&lt;0.001 (p&lt;0.05 for day 23 RENCA data point) for both tumors tested. E) Average tumor size of LLC (left) and RENCA (right) in immunodeficient mice (n=8) treated with 30 mg/kg KVS0001 or vehicle control. Error bars show 95% confidence intervals in all plots.

We then implanted LLC and RENCA cancer cells in the mammary fat pad of C57BL/6N and BALB/c mice respectively and treated with KVS0001 or vehicle control (Figure 5C). The dose of KVS0001 was based on experiments showing that the maximum solubility limit was reached around 2 to 3mg/mL, leading to a maximum single dose of 30mg/kg per mouse per treatment. At this dose, the only toxicity noted was transient weight loss (Figure S20), and no other pathology was observed. Both tumor types experienced significant slowing of tumor growth (Figure 5D). However, when the same tumors were implanted in immunocompromised mice, there was no statistically significant difference in tumor growth between mice treated with KVS0001 or vehicle control (Figure 5E).

We evaluated a total of seven syngeneic mouse cancer cell lines. Four contained a relatively high number of truncating indel mutations, while the remaining three contained a moderate or low number.

None of the three models with moderate or low numbers of truncating mutations responded to KVS0001 (EMT6, B16-F10, M3 Melanoma) (Figure S21). Of the four models with a higher number of truncating mutations, two (LLC and RENCA) responded in a statistically significant fashion while the other two failed to reach significance (CT26, MC38) (Figures 5D and S21).

Discussion

We report the design and execution of a high throughput screen to identify small molecule inhibitors of the nonsense-mediated decay pathway. The method employs a ratiometric output, allowing quantitative and controlled determination of RNA expression changes associated with NMD. Unlike previous NMD screens, this assay is directed at NMD as a process rather than as a way to identify compounds for treating specific mutations related to a disease state, such as cystic fibrosis or β- thalassemia26,28,63. The genetically modified isogenic cell lines employed here are uniquely suited for this approach, producing a high signal-to-noise ratio. Follow- up experiments confirmed a low false positive rate in the screen: four of the eight hits re-expressed transcripts from genes with truncating mutations in a dose-responsive manner when re-tested. Note that our assay allows for the detection of NMD inhibitors regardless of effects on protein translation by virtue of the comparison of wild-type to mutant transcripts for each queried gene. This is an important feature, as many previously reported inhibitors of NMD rely on inhibition of protein translation and thus are unlikely to be optimal for specific restoration of NMD- targeted RNA and protein expression33,36,37,64.

The optimal inhibitor of NMD found in our screen was LY3023414, and we were able to show that this compound inhibited the SMG1 protein kinase. SMG1 acts as the gatekeeper of UPF1 activity through its phosphorylation at multiple sites46. Unfortunately, issues likely related to off-target (pan- kinase) toxicity at the doses required for NMD inhibition prevented LY3023414 from being utilized in further in vivo work. While there are previously described SMG1 inhibitors, all have undesirable off- target or biophysical characteristics complicating their use as therapeutic agents26–29,37,54,65. For example, SMG1i-11 was previously identified as a potent and specific SMG1 inhibitor30. However, in our hands, it was insoluble and neither we nor others have been able to formulate it for administration in vivo. Another unrelated compound has shown successful in vivo re-expression of mutant RNA transcripts, but the effect is reported as less than a two-fold change, compared to the three to four-fold demonstrated here by KVS0001, and little work has been done to develop the compound further in subsequent studies66.

In light of these issues, we attempted to improve the biophysical properties of SMG1i-11 through the development of KVS0001. As reported in this study, KVS0001 was a specific inhibitor of SMG1 that was soluble and could easily be administered to mice. Treatment of KVS0001 not only increased levels of transcripts and proteins from genes with truncating mutations, but also increased the levels of corresponding peptide bound-HLA complexes on the cell surfaces. We developed a new bispecific antibody, KVS-B043, which could recognize these pHLA complexes when the truncating mutations were in the TP53 gene and destroy cancer cells harboring such mutations. In syngeneic mouse models, KVS0001 slowed the growth of some tumors containing multiple truncating mutations, but only when the mice were immunocompetent. This dependence on an intact immune system supports the idea that the slowed growth of the treated tumors was due to recognition of neoantigens.

Despite these encouraging data, the potential utility of KVS0001 for therapeutic purposes remains speculative, for several reasons. First, we do not understand why KVS0001 (and other NMD inhibitors) increase the expression of many, but not all, genes with truncating mutations. More research on the biochemistry of NMD will undoubtedly be required for such understanding, and perhaps the tools described here may facilitate that effort. Second, treatment with KVS0001 reduced the growth of some but not all syngeneic tumor types tested in mice. There appeared to be a correlation between the ability of KVS0001 to inhibit growth and the number of truncating mutations in the tumor. This situation is reminiscent of that encountered with immune checkpoint inhibitors, in which there is a positive but imperfect correlation with tumor mutation burden8,12. Undoubtedly, the ability of the immune system to react to neoantigens is related to their quality as well as their quantity, with quality defined as the ability of the neoantigen to be bound to the host’s particular MHC constitution and initiate an immune response67,68. Third, treatment with KVS0001 led to decreased tumor growth in immunocompetent mice, but did not lead to tumor regressions of the type mandated by RECIST criteria in human clinical trials69,70. Whether this failure to cure tumors in mice results from the contrived nature of the models used – injection of large numbers of rapidly growing cancer cells into animals with relatively little time to react prior to their demise – complicates the interpretation of many pre-clinical immunotherapeutic approaches.

Finally, KVS0001, though it was soluble and could be administered to animals for at least a month, was not entirely non-toxic, causing transient weight loss in some of the mice. Previous work has raised concerns that the importance of NMD in development and normal gene expression make it a difficult pathway to safely disrupt21,71–74. The limited impact on body weight or observed physical activity of mice undergoing KVS0001 therapy at therapeutically relevant doses suggests that NMD inhibition may have acceptable toxicity and be tolerated in developed animals. Another example of this lies in the natural genetic variants of NMD components, which convey variability in the efficiency of NMD between humans, and lends itself to supporting that knockdown of this pathway may be tolerable75–77. Studies in mice also support this observation, as post-development knockdown of UPF1 displayed minimal phenotype change62. The broad use of NMD in normal development and growth, coupled with the observations here, suggests future work with inhibitors of this pathway should continue with a close eye for on-target off-tissue toxicity.

Though there is much work to be done, we hope that the tools, approaches, and compounds described here will facilitate that work. Should the administration of KVS0001 or a related compound prove non-toxic and well-tolerated in humans, we end with a speculation about the possibility of KVS0001 being used as a preventative rather than as a therapeutic agent. Its potential use to prevent the onset of symptoms in pediatric syndromes caused by germline truncating mutations is obvious17,78. Less obvious is the potential for it to prophylactically reduce cancer incidence in patients with hereditary non- polyposis colorectal cancer. These patients inherit heterozygous mutations of a mismatch repair gene, and they do not develop tumors until biallelic mutations of that mismatch repair gene are acquired in a rare stem cell during the second or third decade of life. Similar speculations can be made about the potential of KVS0001 to be used to prevent cancer initiation or progression in patients with other inherited mutations in repair genes, or in individuals exposed to high levels of exogenous mutagens.

eLife Assessment

In this study, the authors developed a cell-based screening assay for the identification of small molecule inhibitors of nonsense-mediated decay (NMD). They used it to validate a novel small molecule SMG1 kinase inhibitor that inhibits NMD in cultured cells leading to the expression of neoantigens from NMD-targeted genes, and in vivo slows tumor growth of cells with a significant number of out-of-frame indel mutations. The conclusions are supported by convincing evidence, and the significance of this work consists in the development of a novel and very promising NMD inhibitor drug that acts as an inhibitor of the SMG1 NMD kinase and is suitable for use in animals. This is an important advance for the field, as previous NMD inhibitors were not specific, lacked efficacy, or were very toxic and hence not suitable for animal application.

Reviewer #1 (Public Review):

Summary:

This work identified new NMD inhibitors and tested them for cancer treatment, based on the hypothesis that inhibiting NMD could lead to the production of cancer neoantigens from the stabilized mutant mRNAs, thereby enhancing the immune system's ability to recognize and kill cancer cells. Key points of the study include:

• Development of an RNA-seq based method for NMD analysis using mixed isogenic cells that express WT or mutant transcripts of STAG2 and TP53 with engineered truncation mutations.

• Application of this method for a drug screen and identified several potential NMD inhibitors.

• Demonstration that one of the identified compounds, LY3023414, inhibits NMD by targeting the SMG1 protein kinase in the NMD pathway in cultured cells and mouse xenografts.

• Due to the in vivo toxicity observed for LY3023414, the authors developed 11 new SMG1 inhibitors (KVS0001-KVS0011) based on the structures of the known SMG1 inhibitor SMG1i-11 and the SMG1 protein itself.

• Among these, KVS0001 stood out for its high potency, excellent bioavailability, and low toxicity in mice. Treatment with KVS0001 caused NMD inhibition and increased presentation of neoantigens on MHC-I molecules, resulting in the clearance of cancer cells in vitro by co-cultured T cells and cancer xenografts in mice by the immune system.

These findings support the strategy of targeting the NMD pathway for cancer treatment and provide new research tools and potential lead compounds for further exploration.

Strengths:

The RNA-seq-based NMD analysis, using isogenic cell lines with specific NMD-inducing mutations, represents a novel approach for the high-throughput identification of potential NMD modulators or genetic regulators. The effectiveness of this method is exemplified by the identification of a new activity of AKT1/mTOR inhibitor LY3023414 in inhibiting NMD.

The properties of KVS0001 described in the manuscript as a novel SMG1 inhibitor suggest its potential as a lead compound for further testing the NMD-targeting strategies in cancer treatment. Additionally, this compound may serve as a useful research tool.

The results of the in vitro cell killing assay and in vivo xenograft experiments in both immuno-proficient and immune-deficient mice indicate that inhibiting NMD could be a viable therapeutic strategy for certain cancers.

Weaknesses:

The authors did not address the potential effects of NMD/SMG1 inhibitors on RNA splicing. Given that the transcripts of many RNA-binding proteins are natural targets of NMD, inhibiting NMD could significantly alter splicing patterns. This, in turn, might influence the outcomes of the RNA-seq-based method for NMD analysis and result interpretation.

While the RNA-seq-based approach offers several advantages for analyzing NMD, the effects of NMD/SMG1 inhibitors observed through this method should be confirmed using established NMD reporters. This step is crucial to rule out the possibility that mutations in STAG2 or TP53 affect NMD in cells, as well as to address potential clonal variations between different engineered cell lines.

The results from the SMG1/UPF1 knockdown and SMG1i-11 experiments presented in Figure 3 correlate with the effects seen for LY3023414, but they do not conclusively establish SMG1 as the direct target of LY3023414 in NMD inhibition. An epistatic analysis with LY3023414 and SMG1-knockdown is needed.

Reviewer #2 (Public Review):

Summary:

Several publications during the past years provided evidence that NMD protects tumor cells from being recognized by the immune system by suppressing the display of neoantigens, and hence NMD inhibition is emerging as a promising anti-cancer approach. However, the lack of an efficacious and specific small-molecule NMD inhibitor with suitable pharmacological properties is currently a major bottleneck in the development of therapies that rely on NMD inhibition. In this manuscript, the authors describe their screen for identifying NMD inhibitors, which is based on isogenic cell lines that either express wild-type or NMD-sensitive transcript isoforms of p53 and STAG2. Using this setup, they screened a library of 2658 FDA-approved or late-phase clinical trial drugs and had 8 hits. Among them they further characterized LY3023414, showing that it inhibits NMD in cultured cells and in a mouse xenograft model, where it, however, was very toxic. Because LY3023414 was originally developed as a PI3K inhibitor, the authors claim that it inhibits NMD by inhibiting SMG1. While this is most likely true, the authors do not provide experimental evidence for this claim. Instead, they use this statement to switch their attention to another previously developed SMG1 inhibitor (SMG1i-11), of which they design and test several derivatives. Of these derivatives, KVS0001 showed the best pharmacological behavior. It upregulated NMD-sensitive transcripts in cultured cells and the xenograft mouse model and two predicted neoantigens could indeed be detected by mass spectrometry when the respective cells were treated with KVS0001. A bispecific antibody targeting T cells to a specific antigen-HLA complex led to increased IFN-gamma release and killing of cancer cells expressing this antigen-HLA complex when they were treated with KVS0001. Finally, the authors show that renal (RENCA) or lung cancer cells (LLC) were significantly inhibited in tumor growth in immunocompetent mice treated with KVS0001. Overall, this establishes KVS0001 as a novel and promising ant-cancer drug that by inhibiting SMG1 (and therewith NMD) increases the neoantigen production in the cancer cells and reveals them to the body's immune system as "foreign".

Strengths:

The novelty and significance of this work consists in the development of a novel and - judging from the presented data - very promising NMD inhibiting drug that is suitable for applications in animals. This is an important advance for the field, as previous NMD inhibitors were not specific, lacked efficacy, or were very toxic and hence not suitable for animal application. It will be still a long way with many challenges ahead towards an efficacious NMD inhibitor that is safe for use in humans, but KVS0001 appears to be a molecule that bears promise for follow-up studies. In addition, while the idea of inhibiting NMD to trigger neoantigen production in cancer cells and so reveal them to the immune system has been around for quite some time, this work provides ample and compelling support for the feasibility of this approach, at least for tumors with a high mutational burden.

Main weaknesses:

There is a disconnect between the screen and the KVS0001 compound, that they describe and test in the second part of the manuscript since KVS0001 is a derivative of the SMG1 inhibitors developed by Gopalsamy et al. in 2012 and not of the lead compound identified in the screen (LY3023414). Because of high toxicity in the mouse xenograft experiments, the authors did not follow up LY3023414 but instead switched to the published SMG1i-11 drug of Gopalsamy and colleagues, a molecule that is widely used among NMD researchers for NMD inhibition in cultured cells. Therefore, in my view, the description of the screen is obsolete, and the paper could just start with the optimization of the pharmacological properties of SMG1i-11 and the characterization of KVS0001. Even though the screen is based on an elegant setup and was executed successfully, it was ultimately a failure as it didn't reveal a useful lead compound that could be further optimized.

Additional points:

- Compared to SMG1i-11, KVS0001 seems less potent in inhibiting SMG1 (higher IC50). It would therefore be important to also compare the specificity of both drugs for SMG1 over other kinases at the applied concentrations (1 uM for SMG1i-11, 5 uM for KVS0001). The Kinativ Assay (Fig. S13) was performed with 100 nM KVS0001, which is 50-fold less than the concentration used for functional assays and hence not really meaningful. In addition, more information on the pharmacokinetic properties and toxicology of KVS0001 would allow a better judgment of the potential of this molecule as a future therapeutic agent.

- On many figures, the concentrations of the used drugs are missing. Please ensure that for every experiment that includes drugs, the drug concentration is indicated.

- Do the authors have an explanation for why LY3023414 has a much stronger effect on the p53 than on the STAG2 nonsense allele (Figure 1B, S8), whereas emetine upregulates the STAG2 nonsense alleles more than the p53 nonsense allele (Figure S5). I find this curious, but the authors do not comment on it.

- While it is a strength of the study that the NMD inhibitors were validated on many different truncation mutations in different cell lines, it would help readers if a table or graphic illustration was included that gives an overview of all mutant alleles tested in this study (which gene, type of mutation, in which cell type). In the current version, this information is scattered throughout the manuscript.

- Lines 194 and 302: That SMG1i-11 was highly insoluble in the hands of the authors is surprising. It is unclear why they used variant 11j, since variant 11e of this inhibitor is widely used among NMD researchers and readily dissolves in DMSO.

- Line 296: The authors claim that they were able to show that LY3023414 inhibited the SMG1 kinase, which is not true. To show this, they would have for example to show that LY3023414 prevents SMG1-mediated UPF1 phosphorylation, as they did for KVS0001 and SMG1i-11 in Fig. 3F. Unless the authors provide this data, the statement should be deleted or modified.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

This work identified new NMD inhibitors and tested them for cancer treatment, based on the hypothesis that inhibiting NMD could lead to the production of cancer neoantigens from the stabilized mutant mRNAs, thereby enhancing the immune system's ability to recognize and kill cancer cells. Key points of the study include:

• Development of an RNA-seq based method for NMD analysis using mixed isogenic cells that express WT or mutant transcripts of STAG2 and TP53 with engineered truncation mutations.

• Application of this method for a drug screen and identified several potential NMD inhibitors.

• Demonstration that one of the identified compounds, LY3023414, inhibits NMD by targeting the SMG1 protein kinase in the NMD pathway in cultured cells and mouse xenografts.

• Due to the in vivo toxicity observed for LY3023414, the authors developed 11 new SMG1 inhibitors (KVS0001-KVS0011) based on the structures of the known SMG1 inhibitor SMG1i-11 and the SMG1 protein itself.

• Among these, KVS0001 stood out for its high potency, excellent bioavailability, and low toxicity in mice. Treatment with KVS0001 caused NMD inhibition and increased presentation of neoantigens on MHC-I molecules, resulting in the clearance of cancer cells in vitro by co-cultured T cells and cancer xenografts in mice by the immune system.

These findings support the strategy of targeting the NMD pathway for cancer treatment and provide new research tools and potential lead compounds for further exploration.

Strengths:

The RNA-seq-based NMD analysis, using isogenic cell lines with specific NMD-inducing mutations, represents a novel approach for the high-throughput identification of potential NMD modulators or genetic regulators. The effectiveness of this method is exemplified by the identification of a new activity of AKT1/mTOR inhibitor LY3023414 in inhibiting NMD.

The properties of KVS0001 described in the manuscript as a novel SMG1 inhibitor suggest its potential as a lead compound for further testing the NMD-targeting strategies in cancer treatment. Additionally, this compound may serve as a useful research tool.

The results of the in vitro cell killing assay and in vivo xenograft experiments in both immuno-proficient and immune-deficient mice indicate that inhibiting NMD could be a viable therapeutic strategy for certain cancers.

Weaknesses:

The authors did not address the potential effects of NMD/SMG1 inhibitors on RNA splicing. Given that the transcripts of many RNA-binding proteins are natural targets of NMD, inhibiting NMD could significantly alter splicing patterns. This, in turn, might influence the outcomes of the RNA-seq-based method for NMD analysis and result interpretation.

This is a very important comment that highlights an important aspect of NMD and potential exciting downstream studies. We did not systematically assess RNA splicing in our work as we are not sure if inhibition of NMD would induce cancer specific splicing that would allow for tumor targeting. It is well established that NMD can impact splicing, including modulating cryptic exon expression, but finding and assessing antigenicity of targetable tumor specific antigens constitutes a study in and of its own. Our own data in figure 4C-F supports this, as a point mutation near a splice site in TP53 strongly induced NMD which was subsequently stopped by KVS0001 treatment. Doing a systematic review of this effect we feel is outside the scope of this manuscript. We’ve incorporated a comment into our discussion highlighting this deficiency, but certainly find the idea of mining RNA-splicing changes an exciting next endeavor.

While the RNA-seq-based approach offers several advantages for analyzing NMD, the effects of NMD/SMG1 inhibitors observed through this method should be confirmed using established NMD reporters. This step is crucial to rule out the possibility that mutations in STAG2 or TP53 affect NMD in cells, as well as to address potential clonal variations between different engineered cell lines.

This is possible, but we want to highlight that all hits from the screen were confirmed in a separate cell line with different clones. While this will not rule out effects to NMD due to STAG2 and TP53 knockdown, the final lead compound was also tested on different endogenous transcripts in both indel and normal transcripts controlled by NMD (i.e., ATF4) in multiple species (human and mouse). Importantly, many of these assays employed the non-mutated transcripts from heterozygous mutant cells to ensure that cis-acting NMD was being measured and to control for any trans-acting splicing or other unanticipated biochemical effects.

The results from the SMG1/UPF1 knockdown and SMG1i-11 experiments presented in Figure 3 correlate with the effects seen for LY3023414, but they do not conclusively establish SMG1 as the direct target of LY3023414 in NMD inhibition. An epistatic analysis with LY3023414 and SMG1-knockdown is needed.

This is a great comment, and is supported by the recent push to confirm drug targets by chemical probes or knockout followed by loss of further effect due to the application of the drug in question. We attempted to knockout SMG1 in multiple cells lines used in this study, including RPE1, MCF10A, NCI-H358 and LS180, and were unable to obtain clones that have biallelic out of frame indels. We were able to obtain multiple clones with in frame indels. Based on our results and those in the publicly available database DepMap we suspect this gene is likely essential, making a simple knockout unfeasible. While this uncertainty is important to keep in mind, we feel it does not detract from the reporting of a novel NMD screen that is mechanistically agnostic and of a novel in vivo active NMD inhibitor.

Reviewer #2 (Public Review):

Summary:

Several publications during the past years provided evidence that NMD protects tumor cells from being recognized by the immune system by suppressing the display of neoantigens, and hence NMD inhibition is emerging as a promising anti-cancer approach. However, the lack of an efficacious and specific small-molecule NMD inhibitor with suitable pharmacological properties is currently a major bottleneck in the development of therapies that rely on NMD inhibition. In this manuscript, the authors describe their screen for identifying NMD inhibitors, which is based on isogenic cell lines that either express wild-type or NMD-sensitive transcript isoforms of p53 and STAG2. Using this setup, they screened a library of 2658 FDA-approved or late-phase clinical trial drugs and had 8 hits. Among them they further characterized LY3023414, showing that it inhibits NMD in cultured cells and in a mouse xenograft model, where it, however, was very toxic. Because LY3023414 was originally developed as a PI3K inhibitor, the authors claim that it inhibits NMD by inhibiting SMG1. While this is most likely true, the authors do not provide experimental evidence for this claim. Instead, they use this statement to switch their attention to another previously developed SMG1 inhibitor (SMG1i-11), of which they design and test several derivatives. Of these derivatives, KVS0001 showed the best pharmacological behavior. It upregulated NMD-sensitive transcripts in cultured cells and the xenograft mouse model and two predicted neoantigens could indeed be detected by mass spectrometry when the respective cells were treated with KVS0001. A bispecific antibody targeting T cells to a specific antigen-HLA complex led to increased IFN-gamma release and killing of cancer cells expressing this antigen-HLA complex when they were treated with KVS0001. Finally, the authors show that renal (RENCA) or lung cancer cells (LLC) were significantly inhibited in tumor growth in immunocompetent mice treated with KVS0001. Overall, this establishes KVS0001 as a novel and promising ant-cancer drug that by inhibiting SMG1 (and therewith NMD) increases the neoantigen production in the cancer cells and reveals them to the body's immune system as "foreign".

Strengths:

The novelty and significance of this work consists in the development of a novel and - judging from the presented data - very promising NMD inhibiting drug that is suitable for applications in animals. This is an important advance for the field, as previous NMD inhibitors were not specific, lacked efficacy, or were very toxic and hence not suitable for animal application. It will be still a long way with many challenges ahead towards an efficacious NMD inhibitor that is safe for use in humans, but KVS0001 appears to be a molecule that bears promise for follow-up studies. In addition, while the idea of inhibiting NMD to trigger neoantigen production in cancer cells and so reveal them to the immune system has been around for quite some time, this work provides ample and compelling support for the feasibility of this approach, at least for tumors with a high mutational burden.

Main weaknesses:

There is a disconnect between the screen and the KVS0001 compound, that they describe and test in the second part of the manuscript since KVS0001 is a derivative of the SMG1 inhibitors developed by Gopalsamy et al. in 2012 and not of the lead compound identified in the screen (LY3023414). Because of high toxicity in the mouse xenograft experiments, the authors did not follow up LY3023414 but instead switched to the published SMG1i-11 drug of Gopalsamy and colleagues, a molecule that is widely used among NMD researchers for NMD inhibition in cultured cells. Therefore, in my view, the description of the screen is obsolete, and the paper could just start with the optimization of the pharmacological properties of SMG1i-11 and the characterization of KVS0001. Even though the screen is based on an elegant setup and was executed successfully, it was ultimately a failure as it didn't reveal a useful lead compound that could be further optimized.

This is a helpful observation from an outside perspective. From our point of view, we were only alerted to the targeting SMG1 due to the previously reported off-target effects of LY3023414 on SMG and lack of plausible explanation for PIK3CA inhibition to efficiently inhibit NMD. We do feel that the screen is worth including for two reasons. First, it offers an unbiased approach for querying the entire NMD pathway for vulnerabilities useful to target. The library chosen was quite small, so the screen itself could be useful to others with larger libraries to test. Second, it did help identify SMG1 as the ideal target for NMD disruption. While targeting SMG1 is not novel, we felt it highlighted why we chose to develop KVS0001. To address this reviewer’s comment, we’ve included a couple sentences in the results and discussion strengthening the point that the screen provided an unbiased approach to finding the best target in the pathway to disrupt NMD and elaborating on the transition from LY3023414 and the screen to development of KVS0001.

Additional points:

- Compared to SMG1i-11, KVS0001 seems less potent in inhibiting SMG1 (higher IC50). It would therefore be important to also compare the specificity of both drugs for SMG1 over other kinases at the applied concentrations (1 uM for SMG1i-11, 5 uM for KVS0001). The Kinativ Assay (Fig. S13) was performed with 100 nM KVS0001, which is 50-fold less than the concentration used for functional assays and hence not really meaningful. In addition, more information on the pharmacokinetic properties and toxicology of KVS0001 would allow a better judgment of the potential of this molecule as a future therapeutic agent.

We agree that the Kinativ assay may have poorly represented the activity of KVS0001 at the bioactive concentration. We have now added 1uM Kinativ data, the highest concentration we were able to run to figure S13.

- On many figures, the concentrations of the used drugs are missing. Please ensure that for every experiment that includes drugs, the drug concentration is indicated.

We apologize for this oversight and have added all drug concentrations on the appropriate plots.

- Do the authors have an explanation for why LY3023414 has a much stronger effect on the p53 than on the STAG2 nonsense allele (Figure 1B, S8), whereas emetine upregulates the STAG2 nonsense alleles more than the p53 nonsense allele (Figure S5). I find this curious, but the authors do not comment on it.

This is an interesting observation. The short answer is we’re not sure. The speculative answer is that it is related to the distinctly different mechanisms of actions of the two inhibitors (see comments from reviewing editor below).

- While it is a strength of the study that the NMD inhibitors were validated on many different truncation mutations in different cell lines, it would help readers if a table or graphic illustration was included that gives an overview of all mutant alleles tested in this study (which gene, type of mutation, in which cell type). In the current version, this information is scattered throughout the manuscript.

This is an excellent suggestion. We’ve included a new table S1 which incorporates the details of each cell line and the genes used in each for this study.

- Lines 194 and 302: That SMG1i-11 was highly insoluble in the hands of the authors is surprising. It is unclear why they used variant 11j, since variant 11e of this inhibitor is widely used among NMD researchers and readily dissolves in DMSO.

As this referee notes SMG1i-11 is soluble in DMSO in our hands as well, which enabled us to use it for our in vitro work. Unfortunately, the concentrations of DMSO required to dissolve the compound to suitable concentrations for in vivo work were too high to safely use in mice with our animal protocols. We also attempted to use ethanol, which also did dissolve SMG1i-11, but led to a significant amount of toxicity in both the drug and vehicle control arms.

- Line 296: The authors claim that they were able to show that LY3023414 inhibited the SMG1 kinase, which is not true. To show this, they would have for example to show that LY3023414 prevents SMG1-mediated UPF1 phosphorylation, as they did for KVS0001 and SMG1i-11 in Fig. 3F. Unless the authors provide this data, the statement should be deleted or modified.

We’ve modified this statement as requested by the referee, now saying we suspected SMG1 was the target based on previously published work.

Recommendations for the authors:

Reviewing Editor (Recommendations For The Authors):

Your paper has been assessed by two reviewers with expertise in the NMD field. They both find the identification and characterization of a new potent and selective inhibitor of the SMG1 NMD kinase with in vivo activity to represent a significant advance in the field, and one that could ultimately be of value as the basis for a novel cancer therapy. However, as you will see both reviewers have concerns about whether the SMG1 inhibitor screen you developed belongs in the paper because it was not used to identify the KVS0001 inhibitor, which instead was generated based on a previously published set of SMG1 inhibitors, and because the NMD inhibitor that did emerge from your screen, LY3023414, was not shown to be a direct inhibitor of SMG1 kinase activity. While it is an elegant screen, during the revision of the paper you could consider streamlining the manuscript by emphasizing how the screening assay was used to validate KVS0001, and bolstering the characterization of the new KVS0001 NMD inhibitor by conducting the proposed additional experiments.

Each of the reviewers raises additional points that should be addressed in a revised version.

The reviewing editor has two additional points:

(1) While emetine inhibits NMD, it is not really a direct NMD inhibitor, as implied, but rather a potent protein synthesis elongation inhibitor that acts by binding to the E-site of the 40S ribosomal subunit, and is therefore, like anisomycin, another protein synthesis inhibitor, working indirectly to inhibit NMD. This should be acknowledged in the section where emetine is first used as an "NMD inhibitor".

This has been included in the indicated section at the referee’s request.

(2) To establish that the observed phenotypic effects of KVS0001 are due to on-target inhibition of SMG1, the authors could generate and express an SMG1 point mutant that is resistant to KVS0001 inhibition, which could be based on the SMG1 catalytic domain structure that the authors used originally to design KVS001. Inhibitor-resistant kinase mutants are the gold standard for demonstrating that the biological consequences of a novel protein kinase inhibitor are due to on-target effects. Admittedly, because SMG1 is such a huge protein, this may be technically challenging and is likely beyond the scope of the present paper.

-We agree with the reviewing editor on all accounts: this would be an ideal experiment to run, but also that it is beyond the scope of the present paper. As indicated in our discussion above with reviewer 1, SMG1 knockout was not possible in our hands, and we suspect it may be due to the gene being essential. Creating an inhibitor resistant mutant could overcome this issue and create an ideal model to test the target for KVS0001. Unfortunately finding such a mutant would likely require significant amounts of trial and error to create a resistant mutant that did not lose SMG1 function. And SMG1 is huge, creating technical issues for experimenting. Due to the anticipated amount of work for such a study we believe this would be better accomplished in future studies.

Reviewer #1 (Recommendations For The Authors):

(1) The authors did not mention a new SMG1 inhibitor and its effects described in Cheruiyot et al, Cancer Res 2019 (PMID: 34215620).

A comment regarding this discovery and its implications for our work was added to the discussion.

(2) There is an inconsistency between the manuscript text and methods sections regarding the time of drug treatment (16 hours vs 14 hours) in the HTS screen.

This has been double checked in our notebook and fixed to reflect 16hrs as the correct incubation time. Thank you for identifying that clerical oversight.

Reviewer #2 (Recommendations For The Authors):

(1) Line 61: The references to NMD reviews are very old (refs 20 and 21). I suggest citing more recent, up-to-date reviews instead.

Two additional references, one from 2016 and another from 2023, have been added to increase support for this statement in the introduction.

(2) Figure S1: Shouldn't the caption of the right panel (TP32 data) say "clone 221" rather than "clone 22"?

This has been fixed.

(3) Figure S18: Please indicate on the y-axis that you are displaying RPKM for p53.

This has been fixed.

(4) Figures 4D and S19: Please indicate concentrations used for all drugs.

This has been fixed.

Version 2

Abstract

Summary

Despite exciting developments in cancer immunotherapy, its broad application is limited by the paucity of targetable antigens on the tumor cell surface. As an intrinsic cellular pathway, nonsense-mediated decay (NMD) conceals neoantigens through the destruction of the RNA products from genes harboring truncating mutations. We developed and conducted a high throughput screen, based on the ratiometric analysis of transcripts, to identify critical mediators of NMD. This screen implicated disruption of kinase SMG1’s phosphorylation of UPF1 as a potential disruptor of NMD. This led us to design a novel SMG1 inhibitor, KVS0001, that elevates the expression of transcripts and proteins resulting from truncating mutations in vivo and in vitro. Most importantly, KVS0001 concomitantly increased the presentation of immune-targetable HLA class I-associated peptides from NMD-downregulated proteins on the surface of cancer cells. KVS0001 provides new opportunities for studying NMD and the diseases in which NMD plays a role, including cancer and inherited diseases.

Introduction

Despite success with cancer immunotherapies, approved immunotherapies are not available for the majority of cancer patients and only a minority of treated patients realize a durable response[1-4]. Current studies largely focus on discovering new agents and identifying patients most likely to benefit from existing immunotherapies[5]. While many studies correlate tumor insertion and deletion (indel) mutation load with immunotherapeutic response, not all tumors with high indel mutational loads respond to checkpoint inhibitors[6-13].

The typical adult solid tumor contains a median of 54 coding somatic nucleotide variants, many of which have the potential to create novel neoantigens or Mutation Associated NeoAntigens (MANA)[14-16]. Approximately 5% of mutations are insertions/deletions (indels) or splice site changes that alter the open reading frame (ORF) of the transcript[14-16]. This subset of mutations is of particular interest because they can result in proteins and derived peptides that are foreign to a host’s healthy cells, giving rise to neoantigens[17-19]. In normal cells, nonsense-mediated decay (NMD) plays an important role in messenger RNA (mRNA) quality control, as well as normal gene expression[20-23]. In cancer cells, however, NMD may aid immuno-evasion by eliminating RNA transcripts coming from genes that carry truncating mutants[19]. This prevents translation and presentation of peptides from these proteins on MHC class I complexes, rendering them invisible to immune cells[24]. Previous work has shown knockdown of the NMD pathway with siRNA enhances the anti-tumor immune response[18, 25].

Analysis of cell lines with loss of the UPF1 RNA helicase and ATPase gene (UPF1), a key mediator of NMD degradation, also showed increased levels of aberrant transcripts and mutant proteins in the alleles targeted by NMD[26]. This is reminiscent of the pharmacological modulation of splicing, which has also been shown to increase the number of neoantigens present on the cancer cell surface due to similar underlying mechanisms[27]. Despite early studies showing little toxicity with NMD inhibition, there is as of yet no reports of a specific chemical inhibitor of the pathway with good bioavailability[28-32].

In this work we create a cell-based high throughput assay to query the effects of a curated library of small molecules on NMD function. From this screen, we identify a lead compound capable of inhibiting NMD and ascertain its protein target as nonsense mediated mRNA decay associated PI3K related kinase (SMG1), a kinase that is a critical mediator of NMD. As the lead compound produced unacceptable toxicity in animal models, we then designed a specific and bioavailable small molecule inhibitor of SMG1 (KVS0001) that is well tolerated in vivo. We then demonstrate that targeted inhibition of SMG1 by KVS0001 leads to the presentation of novel neoantigens identifiable by T cells leading to tumor growth inhibition in vitro and in vivo.

Results

Development of a high throughput assay to identify NMD inhibitors

To develop an assay to find NMD inhibitors, we identified isogenic cell lines with out-of-frame indel mutations, hereinafter referred to as truncating mutations, targeted by NMD activity (Table S1). We previously reported a panel of non-cancerous cell lines in which 19 common tumor suppressor genes were inactivated using the Clustered Regularly Interspaced Short Palindromic Repeats (CRISPR)-Cas9 system[33]. Through evaluation of this panel, we discovered two genes, Stromal Antigen 2 (STAG2) and Tumor Protein p53 (TP53), which did not express their expected proteins when assessed by western blots (STAG2) or immunohistochemistry (p53) (Figure S1). Because the inactivation of these two genes was the result of frame-shift mutations, we suspected that the absence of the proteins was due to NMD. This suspicion was supported by whole transcriptome RNA-sequencing (Figure S2)[33]. Notably we saw an average decrease of STAG2 RNA transcripts by 20-fold and TP53 by 6-fold relative to their respective wild-type transcripts in the parental cell lines.

We selected two STAG2 knockout clones (clone 2 and 8) and one TP53 knockout clone (clone 221) derived from the Retinal Pigmented Epithelial (RPE1) cell line to design a high-throughput screen (HTS). Treatment with the canonical protein synthesis inhibitor, working indirectly to inhibit NMD, emetine demonstrated up to a 60-fold recovery of mutant RNA transcript expression, establishing this panel as appropriate for HTS (Figure S3)[34].

We then designed a Next Generation Sequencing (NGS) assay for NMD (Figure 1A). We mixed the three cell lines in equal proportions and plated the mixture in 96-wells plates followed by treatment with one compound (drug) per well. We determined NMD inhibition efficacy by comparing the wild-type and truncating mutant transcript expression levels in a ratiometric manner. Specifically, the wild-type sequences of the reciprocally knocked out clone (i.e., wild-type STAG2 sequence from the TP53 knockout clones, and wild-type TP53 sequence from the STAG2 knockout clones) served as internal references, providing a ratiometric assay of mutant to wild-type transcript abundance. This ratiometric assay minimized confounders introduced by nonspecific transcriptional activators or generally toxic agents.

LY3023414 is a small molecule capable of increasing transcription of NMD targets.

A) Schematic of HTS used to identify inhibitors of NMD. Mutant transcripts are represented by a smaller length in the cartoon for illustrative purposes only. All small molecules were tested at 10μM. B) Mutant RNA reads relative to wild-type reads for the top 8 hits from the HTS. The dotted line represents the minimum fraction required to be considered a hit (> 5 standard deviations above DMSO control). Full screen results are presented in Figure S4. C) Targeted RNA sequencing results of isogenic RPTec knockout clones treated with the 8 best hits from the HTS at 10µM. The dotted line represents a relative RNA expression level of 1, equal to that of DMSO treated wells. Data for ceritinib, which did not validate on any line, is presented only in Figure S8. D) TP53 western blot on RPE TP53 224, containing a homozygous TP53 mutation, using the four hit compounds that validated in RPTec isogenic lines at 10μM. E) Western blot analysis of full-length TP53α and isoform TP53β after treatment with two NMD inhibitor lead candidates at 10μM. TP53β (expression known to be controlled by NMD) as well as mutant TP53 are prominently induced by LY3023414 whereas full-length is not. Note that RPE TP53 223 is a heterozygous knockout clone with one near wild-type allele whereas RPTec TP53 588 contains a homozygous TP53 indel mutation. F) qPCR showing 10μM LY3023414 treatment causes increased expression of the NMD controlled alternative transcript for TP53, TP53β, in parent cell lines for RPE1 and RPTec. Significance determined by Student’s t-test. Unless indicated otherwise cells were exposed to test compound for 16 hours.

The use of three cell lines with different truncating mutations from two different target genes minimized the possibility that drugs identified in the HTS were cell-line clone or mutation-specific. Note that the use of these cell lines, carefully mixed, banked, and preserved, did not substantially increase the amount of time or work required to screen a single cell line. A combination of well and plate barcodes allowed the pooling and scoring of over 1920 assays in a single NGS lane (see methods).

Execution of a high throughput screen to identify NMD inhibitors

Previous human clinical trials suggest that off-target toxicity at doses required for NMD inhibition makes emetine and other well-known NMD inhibitors unsuitable for human use[35-39]. We performed a high throughput screen (HTS) to identify more specific NMD inhibitors by treating the isogenic cell line panel described above with a commercially available library consisting of 2,658 FDA-approved or in late-phase clinical trial small molecules and natural products (Table S2). After purifying RNA from cells 16 hours post-treatment, we scored NMD inhibition using the strategy described in Figure 1A and Figure S4. Predictably, emetine increased the relative expression of mutant to wild-type transcripts by 3 to 4 fold on average (Figure S5). Eight compounds (0.3% of the library) increased the ratiometric mutant transcript fraction more than five standard deviations above the DMSO controls in all three cell lines (Figure 1B). This hit threshold was chosen as it was the minimum required to ensure no false positives were observed in the DMSO controls. One of these eight compounds, anisomycin, is a known inhibitor of protein synthesis and commonly used NMD inhibitor for in vitro studies, providing independent validation of the screen[34]. The other seven compounds increased mutant RNA transcript levels 5 to 10-fold relative to untreated cells.

LY3023414 inhibits NMD and causes re-expression of mutant RNA and protein

To validate the eight hit compounds described above, we tested their effects in additional lines with mutations targeted by NMD (Table S1). First, we assessed them on isogenic STAG2 and TP53 knockouts in RPtec cells, another non-cancerous cell line (Figures S6A, S6B, and S7)[33]. Four of the original eight hit compounds increased mutant RNA expression in a dose-dependent manner (Figures 1C and S8). Next, we examined the effects of these four compounds on additional RPE1 TP53 knockout cell lines with different mutations predicted to generate truncated TP53 proteins (Figures S6C and S7). While treatment with all four potential NMD inhibitors restored expression of the truncated mutant TP53 proteins, two of them (LY3023414 and halofuginone) did so most robustly (Figures 1D and 1E).

Additionally, TP53 has an isoform, TP53β, whose expression is known to be controlled by the NMD pathway[40]. Using cell lines with intact TP53β isoform transcripts, we observed an increase in the TP53β isoform in cell lines treated with both LY3023414 and halofuginone (Figure 1E middle and left column). Quantitative real time PCR (qPCR) of both the full-length TP53 (TP53α) and NMD sensitive (TP53β) transcripts in parental RPE1 and RPTec cells treated with LY3023414 or halofuginone confirmed the selective upregulation of TP53β, but not TP53α, transcripts (Figure 1F). Based on a consistently stronger effect of LY3023414 over halofuginone across multiple isogenic cell lines and assays, LY3023414 was chosen to be the initial lead compound for further studies involving NMD inhibition.

LY3023414 increases expression of NMD repressed mutant RNA transcripts and proteins in vitro and in vivo

To evaluate whether LY3023414 could relieve NMD repression of naturally occurring heterozygous mutant transcripts, we chose the NCI-H358 and LS180 cancer cell lines (Table S1). Treatment with LY3023414 followed by whole transcriptome RNA sequencing revealed increased expression of the mutant allele in 42% and 67% of heterozygous, out-of-frame, indel mutations in these two lines (Figure 2A). A third of the nonsense mutations in NCI-H358 were also “recovered” (i.e., mutant transcripts increased relative to wildtype transcripts) after treatment with LY3023414 (Figure 2A middle). In LS180 there was only one nonsense mutation meeting the required coverage, so it could not be evaluated in depth. Single base pair substitutions not resulting in stop codons were not affected by LY3023414 in either line (Figure 2A right).

Inhibiting NMD in cancer cells increases broad expression of truncated gene mRNA and protein.

A) Mutant transcript recovery rates for genes containing heterozygous indel mutations based on RNA-sequencing results in cell lines treated with 5µM LY3023414 for 16 hours. Strict inclusion criteria were used, such that only mutations with sufficient sequencing coverage are shown (see methods). Recovery is defined as at least 2-fold increase over DMSO treatment. B) Targeted high coverage RNA-sequencing confirms recovery of mutant transcript levels in NCI-H358 and LS180 cancer cell lines treated with 5µM LY3023414. RNF43 and DROSHA contain common heterozygous SNPs and the mutant allele refers to the non-reference genome allele. Error bars indicate 95% confidence limits. C) Western blot analyses of NCI-H358 cells showing mutant and wild-type protein levels in EXOC1 and D) SPTAN1 with and without 5µM LY3023414 treatment. The black arrow indicates the expected size of the mutant protein. The C-terminal SPTAN1 antibody is downstream of the out-of-frame indel mutation and is not expected to identify the mutant allele. E) Fold change in the number of mutant RNA transcripts from deep-targeted RNA sequencing of heterozygous mutated genes in NCI-H358 and LS180 xenografts treated by oral gavage with 60mg/kg LY3023414 assayed 16 hours post-treatment. Student’s t-test for target genes are all p < 0.05, while the null hypothesis holds for RNF43 (common SNP).

The whole transcriptome sequencing data was confirmed by targeted deep RNA sequencing of two naturally occurring heterozygous truncating mutations from each cell line. Increased expression of the truncated mutant alleles relative to the wild-type alleles of the EXOC1 and SPTAN1 genes was observed in NCI-H358 cells after treatment with LY3023414 (Figure 2B left). Likewise, increased expression of HNRNPA3 and LRPPRC mutant transcripts was observed in LS180 after LY3023414 treatment (Figure 2B right). We noted no significant effects on two coding region heterozygous single nucleotide polymorphisms (SNP’s) after LY3023414 treatment in the genes RNF43 or DROSHA in either cell line (Figure 2B). To substantiate these transcriptomic effects at the protein level, we treated NCI-H358 cells for 24 hours with LY3023414. Western blotting revealed truncated mutant protein for both EXOC1 and SPTAN1 after treatment with LY3023414, whereas the full-length protein levels remained present and unchanged in both the treated and control (DMSO) samples (Figure 2C and 2D left)[41]. An antibody that recognizes the C-terminus of SPTAN1, encoded downstream of the frameshift mutation, did not detect mutant protein, as expected (Figure 2D right).

LY3023414 was originally developed as a phosphatidylinositol 3-kinase (PI3K) inhibitor with activity against AKT Serine/Threonine Kinase 1 (AKT1) and Mammalian Target of Rapamycin (mTOR)[42]. It was tested in a number of clinical trials and has a well-known pharmacokinetic profile in vivo[43-47]. To determine whether LY3023414 affects NMD in vivo, we established xenograft tumors of both NCI-H358 and LS180 in nude mice. Treatment of these mice with a single oral dose of 60 mg/kg LY3023414 led to a significant increase in the expression of mutant RNA transcripts relative to wild-type transcripts 16 hours later (Figure 2E). The RNF43 gene, which harbors a coding region heterozygous SNP, served as a control (Figure 2E). Severe drug-associated toxicity, including weight loss bordering on cachexia and near total inactivity, precluded the longer-term dosing required for anti-tumorigenic effects of LY3023414 in both BALB/c and C57BL/6N immunocompetent mouse strains.

The kinase SMG1 is the target for NMD inhibition by LY3023414

To investigate the mechanism of NMD inhibition by LY3023414, we evaluated the six kinases with the highest reported inhibition by LY3023414[42]. siRNA-mediated knockdown of each of these kinases in NCI-H358 and LS180 cancer cell lines was performed for this purpose (Figure S9). Only knockdown of SMG1 resulted in significant changes in the amount of truncating mutant transcript relative to wild-type transcript in all four genes evaluated (Figure 3A top). RNF43 and DROSHA contain heterozygous coding region SNP’s and served as controls in these experiments and as expected showed no changes despite siRNA treatment (Figure 3A bottom).

Novel NMD inhibitor KVS0001 is SMG1 specific and induces expression of NMD targeted genes in vitro and in vivo.

A) Fraction of mutant allele transcripts in genes with heterozygous indels previously established in this study as sensitive to NMD inhibition. Results show mutant levels after siRNA treatment targeting kinases inhibited by LY3023414. RNF43 and DROSHA are common heterozygous SNPs (shaded gray) and serve as negative controls. B) Fraction of mutant allele transcripts in genes with truncating mutations known to be sensitive to NMD inhibition after siRNA treatment with siUPF1 or non-targeting siRNA. Data from deep-targeted RNA-sequencing. C) Structure of novel NMD inhibitor KVS0001. D) Targeted RNA-sequencing on three genes with heterozygous, out-of-frame, indel mutations in LS180 cancer cells treated in a dose-response with KVS0001 or SMG1i-11. RNF43 serves as a control (common heterozygous SNP) and the mutant allele refers to the non-reference genome allele. E) Western blot of EXOC1 protein in NCI-H358 cells treated with 5µM novel inhibitor KVS0001, LY3023414 or SMG1i-11 for 24 hours. F) Western blot of phosphorylated UPF1 on three cell lines treated with 5µM KVS0001, SMG1i-11, or DMSO. Note that total UPF1 and p-UPF1 were run on different gels, loading controls correspond to indicated gel. G) Fold change in the number of mutant allele transcripts measured by targeted RNA-Seq in genes containing heterozygous out-of-frame indel mutations in NCI-H358 or H) LS180 subcutaneous xenografts in bilateral flanks of nude mice. Mice were treated once with IP injection of vehicle or 30mg/kg KVS0001 and tumors harvested 16hrs post IP treatment. All genes shown contain heterozygous out-of-frame truncating mutations except RNF43 and DROSHA which serve as controls (contain heterozygous SNP’s).

SMG1 is known to regulate the NMD pathway by activating UPF1, an enzyme with RNA helicase and ATPase activity[48, 49]. We therefore knocked down UPF1 with siRNA and found that it restored expression of the NMD-downregulated transcripts, at levels similar to those observed after the knock-down of SMG1 (Figure 3B). Additionally, treatment of NCI-H358 and LS180 cancer cell lines with a previously reported SMG1 specific small molecule inhibitor, SMG1i-11 resulted in specific increases in the transcripts and proteins from genes with truncating mutations, just as did LY3023414 (Figures S10 and S11)[32]. Although SMG1i-11 displayed considerable SMG1 specificity and marked inhibition of NMD, it was highly insoluble. While we had no difficulty getting it into solution in DMSO for in vitro work, we were unable to find a vehicle to administer it in vivo, despite numerous attempts at various formulations and administration routes. This may explain why SMG1i-11 has been demonstrated to be an effective SMG1 inhibitor in vitro, but no peer-reviewed reports of its in vivo activity have been reported to date[32, 50, 51].

Development of an improved NMD inhibitor targeting SMG1

Although we were unable to secure a viable lead compound, the HTS did identify SMG1 as an ideal target to disrupt the nonsense-mediated decay pathway. We sought to develop a new SMG1 inhibitor based on the cryo-electron microscopy structure of the binding pocket of SMG1[52, 53]. We attempted the synthesis of eleven compounds (KVS0001 to KVS0011) and tested for bioavailability and preservation of target specificity. Among these, KVS0001 stood out due to its solubility while preserving SMG1 inhibitory activity (Table S2, Figures 3C and S12). Mass spectrometry-based assays showed that KVS0001 inhibits SMG1 protein more than any of the other 246 protein or lipid kinases tested at concentrations from 10nM to 1µM (Table S4 and Figure S13)[54]. Noteworthy off-target kinase inhibition was not observed until doses of 1µM and above. Experiments using NCI-H358 and LS180 cells showed that KVS0001 is bioactive in the nanomolar range and subverts the NMD-mediated down-regulation of truncating mutant transcripts and proteins (Figures 3D, 3E, S14, S15). Inhibition at concentrations as low as 600nM led to equal expression of wild-type and mutant transcripts, suggesting near total blockade of the NMD pathway (Figures 3D, S14). Western blotting showed that KVS0001 substantially decreases the amount of phosphorylated UPF1, the downstream target of SMG1 kinase activity, in three different cell lines (Figure 3F). Finally, KVS0001 treatment of NCI-H358 and LS180-derived xenograft tumors in nude mice resulted in significant increases in transcript levels in each of six tested endogenous genes with truncating mutations, while having had no measurable effects on genes containing heterozygous coding region SNP’s (Figures 3G and 3H).

NMD inhibition with KVS0001 causes MHC class I display of hidden neoantigens

Cancer cells may evade immune surveillance by downregulating genes with truncating mutations as a result of NMD[19, 55]. Indeed, previous studies with non-specific or toxic NMD inhibitors have shown an increase in selected antigens from tumor-specific mutations when NMD is inhibited[56]. We therefore investigated whether KVS0001 could similarly alter the cell surface presentation of proteins from genes harboring truncating mutations in NCI-H358 and LS180 cells. Using quantitative HPLC-mass spectrometry (MS) we observed a striking (45 to 90 fold) increase in the EXOC1 and RAB14 derived neoantigens, and a significant (2 fold) increase in the ZDHHC16 derived neoantigen (Figures 4A, 4B, S16, S17)[57]. This is consistent with the re-expression of mutant transcript and protein shown previously in this study (Figures 3D and 3E). These three peptides were chosen based on an in-silico review of potentially presented peptides and a preliminary experiment that looked at qualitative (present or not present) presentation of the predicted peptides[58, 59].

KVS0001 treatment induces targetable cell surface presentation of peptides known to be downregulated by NMD.

A) MHC class I HLA presentation of mutant specific peptide sequences from NCI-H358 and B) LS180 cells by quantitative HPLC-Mass Spectrometry treated with DMSO or 5µM KVS0001. The gene name, type of mutation (in parenthesis) and presented peptide are shown on the y-axis for each gene. Colors indicate different ions. C) TP53 gene structure and mutant DNA sequence for NCI-H716 and NCI-H2228 cancer cell lines, both contain a homozygous splice site mutation in TP53. Capital letters represent exonic sequence; lowercase letters represent intronic sequence. DNA mutation reflected by gold bases. D) Western blot against TP53 in the presence or absence of 5µM NMD inhibitor in NCI-H716_A24 and NCI-H2228 cell lines. NCI-H2228 has an expected size of 46.6 kDa and NCI-H716 of 34.7 kDa. E) IFN-γ levels over baseline based on ELISA in a co-culture assay with NCI-H716_A24 and NCI-H2228 cells, 1.25µM NMD inhibitor, human CD3+ T-cells, and bispecific antibody for TP53 and CD3. Chemotherapy (5-Fluorouracil) is shown as a control. F) Cell killing based on luciferase levels in a co-culture assay in NCI-H716 cells with and without A24 expression, treated with TP53-CD3 bispecific antibody, 1.25µM NMD inhibitor and human CD3+ T-cells.

Targetable peptide presentation occurs with NMD inhibition by KVS0001

To test whether cancer cell neoantigens presented as a result of NMD inhibition could be targeted, we evaluated two cancer cell lines, NCI-H716 and NCI-H2228 (Table S1). Both lines contain homozygous truncating mutations in TP53 which produce transcripts down-regulated by NMD, with each exhibiting RNA levels less than 20% of the median level of expression among 675 cancer cell lines (Figures 4C and S18)[60, 61]. Treatment with KVS0001 increased the expression of TP53 in both lines, while the commonly used therapeutic agents 5-fluorouracil and etoposide did not affect TP53 protein abundance (Figures 4D and S19).

To determine whether this disruption of NMD repression is targetable by T-cells, we developed a bispecific antibody (KVS-BI043) that recognizes a peptide-HLA complex on one end and CD3 on the other end. CD3 is expressed only on T-cells, and this bispecific antibody functions as a T-cell engager, linking target cells to cytotoxic T-cells, which then kill the targets[62, 63]. KVS-BI043 recognizes a ten amino acid peptide (residues 125 to 134 of TP53) bound to HLA-A24. NCI-H2228 naturally express A24 whereas NCI-H716 cells were engineered to express A24 using a retrovirus (NCI-H716_A24). Treatment of NCI-H716_A24 or NCI-H2228 cells with KVS0001 and the KVS-BI043 bispecific antibody in the presence of normal T-cells caused a significant increase in interferon (IFN-γ) release (Figure 4E). We observed no changes in IFN-γ levels in the absence of KVS-BI043 or normal T-cells (Figure 4E grey boxed lanes). Most importantly, treatment of NCI-H716_A24 with KVS0001 in the presence of KVS-BI043 and T-cells led to significant killing of the target cancer cells, which was also not observed in the absence of T-cells or the absence of HLA-24 in the target cells (Figure 4F). NCI-H2228 was assessed for killing but expressed too much TP53 at baseline (Figure 4D left most lane) and thus displayed substantial killing even in the DMSO controls.

Tumor growth is slowed in mice treated with KVS0001

Finally, we investigated whether KVS0001 could impact tumor growth in syngeneic models in which the native immune system might play a role (Table S1). For this purpose, we first used murine RENCA (renal cancer) and LLC (lung cancer) as they are known to have a relatively large number of out-of-frame indel mutations (Table S5). Although murine and human SMG1 are highly related (98% at the amino acid level), it was important to show that KVS0001 could actually inhibit NMD in murine cells.

For this purpose, we tested eight genes in LLC, and four in RENCA, which contained out-of-frame indel mutations potentially targeted by NMD. We also assessed the expression of three genes without any mutations known to have their normal expression controlled by NMD[64]. We found that six of the twelve truncating mutation-containing genes and five of the six expression controlled by NMD genes had significantly increased RNA following in vitro treatment with KVS0001 (Figure 5A and 5B).

In vivo treatment of murine tumors with KVS0001 yield differential tumor growth compared with vehicle treatment.

A) Fold change in RNA transcript levels in LLC or B) RENCA cells treated in-vitro with 5µM of NMD inhibitor KVS0001 or DMSO. Orange bars indicate genes with homozygous indel mutations potentially targeted by NMD. Purple bars show genes with no mutations but that are known to have their normal transcription levels controlled by NMD. Green bar is a control gene that should not change with treatment. The dotted line shows relative expression of DMSO treatment (equal to 1). C) Treatment schedule for syngeneic tumor mouse experiments. D) Average tumor size of LLC (left) and RENCA (right) syngeneic tumors in immune-competent mice (n=8) treated with 30 mg/kg KVS0001 or vehicle control. Difference is statistically significant after day 10 based on one-way ANOVA with Dunnett’s test p<0.001 (p<0.05 for day 23 RENCA data point) for both tumors tested. E) Average tumor size of LLC (left) and RENCA (right) in immunodeficient mice (n=8) treated with 30 mg/kg KVS0001 or vehicle control. Error bars show 95% confidence intervals in all plots.

We then implanted LLC and RENCA cancer cells in the mammary fat pad of C57BL/6N and BALB/c mice respectively and treated with KVS0001 or vehicle control (Figure 5C). The dose of KVS0001 was based on experiments showing that the maximum solubility limit was reached around 2 to 3mg/mL, leading to a maximum single dose of 30mg/kg per mouse per treatment. At this dose, the only toxicity noted was transient weight loss (Figure S20), and no other pathology was observed. Both tumor types experienced significant slowing of tumor growth (Figure 5D). However, when the same tumors were implanted in immunocompromised mice, there was no statistically significant difference in tumor growth between mice treated with KVS0001 or vehicle control (Figure 5E).

We evaluated a total of seven syngeneic mouse cancer cell lines. Four contained a relatively high number of truncating indel mutations, while the remaining three contained a moderate or low number. None of the three models with moderate or low numbers of truncating mutations responded to KVS0001 (EMT6, B16-F10, M3 Melanoma) (Figure S21). Of the four models with a higher number of truncating mutations, two (LLC and RENCA) responded in a statistically significant fashion while the other two failed to reach significance (CT26, MC38) (Figures 5D and S21).

Discussion

We report the design and execution of a high throughput screen to identify small molecule inhibitors of the nonsense-mediated decay pathway. The method employs a ratiometric output, allowing quantitative and controlled determination of RNA expression changes associated with NMD. Unlike previous NMD screens, this assay is directed at NMD as a process rather than as a way to identify compounds for treating specific mutations related to a disease state, such as cystic fibrosis or β-thalassemia[28, 30, 65]. The genetically modified isogenic cell lines employed here are uniquely suited for this approach, producing a high signal-to-noise ratio. Follow-up experiments confirmed a low false positive rate in the screen: four of the eight hits re-expressed transcripts from genes with truncating mutations in a dose-responsive manner when re-tested. Note that our assay allows for the detection of NMD inhibitors regardless of effects on protein translation by virtue of the comparison of wild-type to mutant transcripts for each queried gene. This is an important feature, as many previously reported inhibitors of NMD, such as emetine and anisomycin, rely on inhibition of protein translation and thus are unlikely to be optimal for specific restoration of NMD-targeted RNA and protein expression[35, 38, 39, 66]. During the review process it was also noted that different NMD inhibitors caused re-expression of different knockout clones to different extents (see Figure 1B and S8 as an example). It is not clear to us why this occurred, but we speculate that it may be related to differences in the mechanism of action (i.e., protein synthesis inhibition based NMD effects versus targeted NMD pathway inhibition as one example).

The optimal inhibitor of NMD found in our screen was LY3023414, and we suspected that this compound may inhibit the kinase SMG1 based on previously published kinase activity data with this small molecule[42]. SMG1 acts as the gatekeeper of UPF1 activity through its phosphorylation at multiple sites[48]. Unfortunately, issues likely related to off-target (pan-kinase) toxicity at the doses required for NMD inhibition prevented LY3023414 from being utilized in further in vivo work. While there are previously described SMG1 inhibitors, all have undesirable off-target or biophysical characteristics complicating their use as therapeutic agents[28-31, 39, 56, 67]. For example, SMG1i-11 was previously identified as a potent and specific SMG1 inhibitor[32]. However, in our hands, it was not soluble at concentrations required for in vivo work in a suitable vehicle for mouse administration.

Another unrelated compound has shown successful in vivo re-expression of mutant RNA transcripts, but the effect is reported as less than a two-fold change, compared to the three to four-fold demonstrated here by KVS0001, and little work has been done to develop the compound further in subsequent studies[68].

Despite these chemistry related challenges, our screen offered an unbiased assessment of the best targets for NMD pathway disruption, and thus we concluded that SMG1 was an ideal protein to inhibit for this purpose. In light of this, we attempted to improve the biophysical properties of SMG1i-11 through the development of KVS0001. As reported in this study, KVS0001 was a specific inhibitor of SMG1 that was soluble and could easily be administered to mice. Treatment of KVS0001 not only increased levels of transcripts and proteins from genes with truncating mutations, but also increased the levels of corresponding peptide bound-HLA complexes on the cell surfaces. We developed a new bispecific antibody, KVS-B043, which could recognize these pHLA complexes when the truncating mutations were in the TP53 gene and destroy cancer cells harboring such mutations. In syngeneic mouse models, KVS0001 slowed the growth of some tumors containing multiple truncating mutations, but only when the mice were immunocompetent. This dependence on an intact immune system supports the idea that the slowed growth of the treated tumors was due to recognition of neoantigens.

Despite these encouraging data, the potential utility of KVS0001 for therapeutic purposes remains speculative, for several reasons. First, we do not understand why KVS0001 (and other NMD inhibitors) increase the expression of many, but not all, genes with truncating mutations. We also do not systematically address the changes to splicing that are undoubtedly occurring when NMD is disrupted in either normal or tumor tissue. More research on the biochemistry of NMD will undoubtedly be required for such understanding, and perhaps the tools described here may facilitate that effort. Second, treatment with KVS0001 reduced the growth of some but not all syngeneic tumor types tested in mice. There appeared to be a correlation between the ability of KVS0001 to inhibit growth and the number of truncating mutations in the tumor. Likewise, recent evidence has been presented showing that cells deficient in known splicing factors are sensitive to subsequent NMD inhibition[69]. This situation is reminiscent of that encountered with immune checkpoint inhibitors, in which there is a positive but imperfect correlation with loss of microsatellite instability (MSI) related proteins and tumor mutation burden[8, 12]. Undoubtedly, the ability of the immune system to react to neoantigens is related to their quality as well as their quantity, with quality defined as the ability of the neoantigen to be bound to the host’s particular MHC constitution and initiate an immune response[70, 71]. Third, treatment with KVS0001 led to decreased tumor growth in immunocompetent mice, but did not lead to tumor regressions of the type mandated by RECIST criteria in human clinical trials[72, 73]. Whether this failure to cure tumors in mice results from the contrived nature of the models used – injection of large numbers of rapidly growing cancer cells into animals with relatively little time to react prior to their demise – complicates the interpretation of many pre-clinical immunotherapeutic approaches.

Finally, KVS0001, though it was soluble and could be administered to animals for at least a month, was not entirely non-toxic, causing transient weight loss in some of the mice. Previous work has raised concerns that the importance of NMD in development and normal gene expression make it a difficult pathway to safely disrupt[21, 74-77]. The limited impact on body weight or observed physical activity of mice undergoing KVS0001 therapy at therapeutically relevant doses suggests that NMD inhibition may have acceptable toxicity and be tolerated in developed animals. Another example of this lies in the natural genetic variants of NMD components, which convey variability in the efficiency of NMD between humans, and lends itself to supporting that knockdown of this pathway may be tolerable[78-80]. Studies in mice also support this observation, as post-development knockdown of UPF1 displayed minimal phenotype change[64]. The broad use of NMD in normal development and growth, coupled with the observations here, suggests future work with inhibitors of this pathway should continue with a close eye for on-target off-tissue toxicity.

Though there is much work to be done, we hope that the tools, approaches, and compounds described here will facilitate that work. Should the administration of KVS0001 or a related compound prove non-toxic and well-tolerated in humans, we end with a speculation about the possibility of KVS0001 being used as a preventative rather than as a therapeutic agent. Its potential use to prevent the onset of symptoms in pediatric syndromes caused by germline truncating mutations is obvious[17, 81]. Less obvious is the potential for it to prophylactically reduce cancer incidence in patients with hereditary non-polyposis colorectal cancer. These patients inherit heterozygous mutations of a mismatch repair gene, and they do not develop tumors until biallelic mutations of that mismatch repair gene are acquired in a rare stem cell during the second or third decade of life. Similar speculations can be made about the potential of KVS0001 to be used to prevent cancer initiation or progression in patients with other inherited mutations in repair genes, or in individuals exposed to high levels of exogenous mutagens.

Supporting information